---
title: "Forecasting Stock Market Volatility Using Daily S&P 500 ETF Returns"
author: "He Jiang, Patrick Zhu"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, message=FALSE, warning=FALSE}
library(quantmod)
library(zoo)
library(xts)
library(readr)
library(dplyr)
library(ggplot2)
library(forecast)
library(tseries)
```

# Introduction

This project studies stock market volatility using daily S&P 500 ETF returns. Instead of forecasting raw stock prices, we focus on volatility because financial returns often display volatility clustering and are more appropriate for time series modeling.

# Data Download and Preparation

We use daily data for the SPDR S&P 500 ETF Trust (SPY) downloaded from Yahoo Finance. Adjusted closing prices are used to compute daily log returns, and volatility is measured as the 21-day rolling standard deviation of returns.

```{r download-data}
# Download SPY data from Yahoo Finance
getSymbols("SPY", src = "yahoo", from = "2015-01-01")

# Extract adjusted close price
spy_adj <- Ad(SPY)

# Compute daily log returns
spy_ret <- diff(log(spy_adj))

# Compute 21-day rolling volatility
spy_vol_21 <- rollapply(
  spy_ret,
  width = 21,
  FUN = sd,
  align = "right",
  fill = NA
)

# Build data frame
data <- data.frame(
  date = as.Date(index(spy_vol_21)),
  adj_close = as.numeric(spy_adj[index(spy_vol_21)]),
  return = as.numeric(spy_ret[index(spy_vol_21)]),
  volatility_21 = as.numeric(spy_vol_21)
)

data <- na.omit(data)

head(data)
```

# Exploratory Data Analysis

## SPY Adjusted Close

```{r plot-price}
ggplot(data, aes(x = date, y = adj_close)) +
  geom_line() +
  labs(title = "SPY Adjusted Closing Price", x = "Date", y = "Adjusted Close") +
  theme_minimal()
```

## Daily Returns

```{r plot-returns}
ggplot(data, aes(x = date, y = return)) +
  geom_line() +
  labs(title = "Daily Log Returns of SPY", x = "Date", y = "Log Return") +
  theme_minimal()
```

## 21-Day Rolling Volatility

```{r plot-volatility}
ggplot(data, aes(x = date, y = volatility_21)) +
  geom_line() +
  labs(title = "21-Day Rolling Volatility of SPY Returns", x = "Date", y = "Volatility") +
  theme_minimal()
```

# Preliminary Time Series Analysis

## Summary Statistics

```{r summary-stats}
summary(data)
```

## ACF and PACF of Volatility

```{r acf-pacf}
vol_ts <- ts(data$volatility_21, frequency = 252)
Acf(vol_ts, main = "ACF of 21-Day Rolling Volatility")
Pacf(vol_ts, main = "PACF of 21-Day Rolling Volatility")
```

# Next Steps

The next stage of the project will split the data into training and test sets and compare forecasting performance across multiple models, such as ARIMA and GARCH.

# Train/Test Split

We split the volatility series into a training set and a test set. The training set is used to fit the models, and the test set is used to evaluate out-of-sample forecasting performance.

```{r train-test-split}
n <- nrow(data)
train_size <- floor(0.8 * n)

train_data <- data[1:train_size, ]
test_data  <- data[(train_size + 1):n, ]

train_vol <- train_data$volatility_21
test_vol  <- test_data$volatility_21

length(train_vol)
length(test_vol)

tail(train_data)
head(test_data)
```

# Baseline Forecast

As a simple benchmark, we first use a naive forecasting approach. This baseline helps us evaluate whether more advanced models such as ARIMA and GARCH actually improve forecasting accuracy.

```{r}
train_ts <- ts(train_vol, start = 1, frequency = 1)
test_ts  <- ts(test_vol,  start = length(train_ts) + 1, frequency = 1)

baseline_fit <- naive(train_ts, h = length(test_ts))

autoplot(baseline_fit) +
  autolayer(test_ts, series = "Actual") +
  labs(
    title = "Baseline Naive Forecast vs Actual Volatility",
    x = "Time",
    y = "Volatility"
  ) +
  theme_minimal()
```




```{r baseline-accuracy}
accuracy(baseline_fit, test_ts)
```


```{r}
arima_fit <- auto.arima(train_ts, seasonal = FALSE)
summary(arima_fit)
```

```{r arima-forecast}
arima_fc <- forecast(arima_fit, h = length(test_ts))

autoplot(arima_fc) +
  autolayer(test_ts, series = "Actual") +
  labs(
    title = "ARIMA Forecast vs Actual Volatility",
    x = "Time",
    y = "Volatility"
  ) +
  theme_minimal()
```
```{r arima-accuracy}
accuracy(arima_fc, test_ts)
```
```{r compare-baseline-arima}
baseline_acc <- accuracy(baseline_fit, test_ts)
arima_acc    <- accuracy(arima_fc, test_ts)

comparison <- data.frame(
  Model = c("Naive Baseline", "ARIMA"),
  RMSE  = c(baseline_acc[2, "RMSE"], arima_acc[2, "RMSE"]),
  MAE   = c(baseline_acc[2, "MAE"],  arima_acc[2, "MAE"]),
  MAPE  = c(baseline_acc[2, "MAPE"], arima_acc[2, "MAPE"])
)

comparison
```

## GARCH Model

```{r garch-model, message=FALSE, warning=FALSE}
library(tseries)

# Use return series for GARCH
train_ret <- train_data$return
test_ret  <- test_data$return

# Fit GARCH(1,1)
garch_fit <- garch(train_ret, order = c(1, 1))

# Show model summary
summary(garch_fit)

# Extract fitted conditional volatility from training period
train_garch_vol <- fitted(garch_fit)[, 1]

# Use the last fitted conditional volatility as a simple forecast benchmark
last_garch_vol <- tail(train_garch_vol, 1)
garch_forecast <- rep(last_garch_vol, length(test_vol))

# Convert forecast to time series for plotting
garch_fc_ts <- ts(garch_forecast, start = length(train_ts) + 1, frequency = 1)

# Plot GARCH forecast against actual test volatility
autoplot(garch_fc_ts, series = "GARCH Forecast") +
  autolayer(test_ts, series = "Actual") +
  labs(
    title = "GARCH-Based Forecast vs Actual Volatility",
    x = "Time",
    y = "Volatility"
  ) +
  theme_minimal()

# Accuracy
garch_acc <- accuracy(garch_fc_ts, test_ts)
garch_acc
```

```{r compare-models}
baseline_acc <- accuracy(baseline_fit, test_ts)
arima_acc    <- accuracy(arima_fc, test_ts)
garch_acc    <- accuracy(garch_fc_ts, test_ts)

comparison <- data.frame(
  Model = c("Naive Baseline", "ARIMA", "GARCH-based"),
  RMSE  = c(baseline_acc[2, "RMSE"], arima_acc[2, "RMSE"], garch_acc[1, "RMSE"]),
  MAE   = c(baseline_acc[2, "MAE"],  arima_acc[2, "MAE"],  garch_acc[1, "MAE"]),
  MAPE  = c(baseline_acc[2, "MAPE"], arima_acc[2, "MAPE"], garch_acc[1, "MAPE"])
)

comparison
```
## ARIMA Residual Diagnostics

```{r arima-residuals}
checkresiduals(arima_fit)
```

## ETS Model

```{r ets-model}
ets_fit <- ets(train_ts)
ets_fc  <- forecast(ets_fit, h = length(test_ts))

autoplot(ets_fc) +
  autolayer(test_ts, series = "Actual") +
  labs(
    title = "ETS Forecast vs Actual Volatility",
    x = "Time",
    y = "Volatility"
  ) +
  theme_minimal()

ets_acc <- accuracy(ets_fc, test_ts)
ets_acc
```


```{r nnar-model}
nn_fit <- nnetar(train_ts)
nn_fc  <- forecast(nn_fit, h = length(test_ts))

autoplot(nn_fc) +
  autolayer(test_ts, series = "Actual") +
  labs(
    title = "NNAR Forecast vs Actual Volatility",
    x = "Time",
    y = "Volatility"
  ) +
  theme_minimal()

nn_acc <- accuracy(nn_fc, test_ts)
nn_acc
```


```{r results-table}

results_all <- data.frame(
  Model = c("Naive", "ARIMA", "GARCH", "ETS", "NNAR"),
  RMSE  = c(
    baseline_acc[2, "RMSE"],
    arima_acc[2, "RMSE"],
    garch_acc[1, "RMSE"],
    ets_acc[2, "RMSE"],
    nn_acc[2, "RMSE"]
  ),
  MAE   = c(
    baseline_acc[2, "MAE"],
    arima_acc[2, "MAE"],
    garch_acc[1, "MAE"],
    ets_acc[2, "MAE"],
    nn_acc[2, "MAE"]
  ),
  MAPE  = c(
    baseline_acc[2, "MAPE"],
    arima_acc[2, "MAPE"],
    garch_acc[1, "MAPE"],
    ets_acc[2, "MAPE"],
    nn_acc[2, "MAPE"]
  )
)

results_long <- data.frame(
  Model = rep(results_all$Model, 3),
  Metric = rep(c("RMSE", "MAE", "MAPE"), each = nrow(results_all)),
  Value = c(results_all$RMSE, results_all$MAE, results_all$MAPE)
)

ggplot(results_long, aes(x = Model, y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  labs(
    title = "Forecast Accuracy Comparison Across Models",
    x = "Model",
    y = "Error Value"
  ) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

```
```{r}
ggplot(results_long, aes(x = Model, y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  labs(
    title = "Forecast Accuracy Comparison Across Models",
    x = "Model",
    y = "Error Value"
  ) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```
```{r}

ggplot(results_long, aes(x = Metric, y = Model, fill = Value)) +
  geom_tile() +
  geom_text(aes(label = round(Value, 4)), size = 4) +
  labs(
    title = "Heatmap of Forecast Errors",
    x = "Metric",
    y = "Model"
  ) +
  theme_minimal()
```

```{r}


results_rank <- data.frame(
  Model = results_all$Model,
  RMSE_rank = rank(results_all$RMSE),
  MAE_rank  = rank(results_all$MAE),
  MAPE_rank = rank(results_all$MAPE)
)


ranking_long <- data.frame(
  Model = rep(results_rank$Model, 3),
  Metric = rep(c("RMSE", "MAE", "MAPE"), each = nrow(results_rank)),
  Rank = c(results_rank$RMSE_rank,
           results_rank$MAE_rank,
           results_rank$MAPE_rank)
)

ggplot(ranking_long, aes(x = Metric, y = Model, fill = Rank)) +
  geom_tile() +
  geom_text(aes(label = Rank), size = 5) +
  labs(
    title = "Model Ranking by Forecast Metric (1 = Best)",
    x = "Metric",
    y = "Model"
  ) +
  theme_minimal()

```

```{r}

forecast_df <- data.frame(
  Time   = 1:length(test_ts),
  Actual = as.numeric(test_ts),
  Naive  = as.numeric(baseline_fit$mean),
  ARIMA  = as.numeric(arima_fc$mean),
  GARCH  = as.numeric(garch_fc_ts),
  ETS    = as.numeric(ets_fc$mean),
  NNAR   = as.numeric(nn_fc$mean)
)

forecast_long <- data.frame(
  Time = rep(forecast_df$Time, 6),
  Series = c(
    rep("Actual", nrow(forecast_df)),
    rep("Naive",  nrow(forecast_df)),
    rep("ARIMA",  nrow(forecast_df)),
    rep("GARCH",  nrow(forecast_df)),
    rep("ETS",    nrow(forecast_df)),
    rep("NNAR",   nrow(forecast_df))
  ),
  Value = c(
    forecast_df$Actual,
    forecast_df$Naive,
    forecast_df$ARIMA,
    forecast_df$GARCH,
    forecast_df$ETS,
    forecast_df$NNAR
  )
)

ggplot(forecast_long, aes(x = Time, y = Value, color = Series)) +
  geom_line() +
  labs(
    title = "Actual vs Forecasted Volatility on the Test Set",
    x = "Time Index",
    y = "Volatility"
  ) +
  theme_minimal()

```




