---
title: "Forecasting Stock Market Volatility Using Daily S&P 500 ETF Returns"
author: "He Jiang, Patrick Zhu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
packages <- c("quantmod", "zoo", "xts", "readr", "dplyr", "ggplot2", "forecast")
installed <- packages %in% rownames(installed.packages())
if (any(!installed)) {
  install.packages(packages[!installed])
}

library(quantmod)
library(zoo)
library(xts)
library(readr)
library(dplyr)
library(ggplot2)
library(forecast)
```

# Introduction

This project studies stock market volatility using daily S&P 500 ETF returns. Instead of forecasting raw stock prices, we focus on volatility because financial returns often display volatility clustering and are more appropriate for time series modeling.

# Data Download and Preparation

We use daily data for the SPDR S&P 500 ETF Trust (SPY) downloaded from Yahoo Finance. Adjusted closing prices are used to compute daily log returns, and volatility is measured as the 21-day rolling standard deviation of returns.

```{r download-data}
# Download SPY data from Yahoo Finance
getSymbols("SPY", src = "yahoo", from = "2015-01-01")

# Extract adjusted close price
spy_adj <- Ad(SPY)

# Compute daily log returns
spy_ret <- diff(log(spy_adj))

# Compute 21-day rolling volatility
spy_vol_21 <- rollapply(
  spy_ret,
  width = 21,
  FUN = sd,
  align = "right",
  fill = NA
)

# Build data frame
data <- data.frame(
  date = as.Date(index(spy_vol_21)),
  adj_close = as.numeric(spy_adj[index(spy_vol_21)]),
  return = as.numeric(spy_ret[index(spy_vol_21)]),
  volatility_21 = as.numeric(spy_vol_21)
)

data <- na.omit(data)

head(data)
```

# Exploratory Data Analysis

## SPY Adjusted Close

```{r plot-price}
ggplot(data, aes(x = date, y = adj_close)) +
  geom_line() +
  labs(title = "SPY Adjusted Closing Price", x = "Date", y = "Adjusted Close") +
  theme_minimal()
```

## Daily Returns

```{r plot-returns}
ggplot(data, aes(x = date, y = return)) +
  geom_line() +
  labs(title = "Daily Log Returns of SPY", x = "Date", y = "Log Return") +
  theme_minimal()
```

## 21-Day Rolling Volatility

```{r plot-volatility}
ggplot(data, aes(x = date, y = volatility_21)) +
  geom_line() +
  labs(title = "21-Day Rolling Volatility of SPY Returns", x = "Date", y = "Volatility") +
  theme_minimal()
```

# Preliminary Time Series Analysis

## Summary Statistics

```{r summary-stats}
summary(data)
```

## ACF and PACF of Volatility

```{r acf-pacf}
vol_ts <- ts(data$volatility_21, frequency = 252)
Acf(vol_ts, main = "ACF of 21-Day Rolling Volatility")
Pacf(vol_ts, main = "PACF of 21-Day Rolling Volatility")
```

# Next Steps

The next stage of the project will split the data into training and test sets and compare forecasting performance across multiple models, such as ARIMA and GARCH.

